---
title: "Beachhead Assignment #1"
format: html
editor: visual
---

## Introduction

I'm using the MTA Subway Hourly Ridership dataset (2020–2024) from data.ny.gov, chosen for its relevance to my work at NYCT and because it needs to be cleaned up before we can draw any conclusions. Ridership will serve as the target variable. Since this dataset is very large at over 500,000 rows, I have queried it so it is much more simple to work and with and publish by limiting it to the month of July 2024 consisting of riders with at least 1 transfer.

To tackle the problem, I will parse the timestamp into a proper datetime, drop redundant columns, split any categorical fields into separate variables, and subset to a clean, well-named set of columns.

The data challenges I anticipate include a string-formatted timestamp that needs to be parsed and separated to increase the ways of sorting and filtering through the data, redundant or overlapping columns, a compound categorical field that conflates two variables when they don't relate, a large row count that may need filtering, and multiple transit modes that will require a scoping decision.

## Code deliverables

First, the necessary libraries must be loaded in order to load the dataset and access it.

```{r}
library(tidyverse)

dataMta <- read_csv("https://raw.githubusercontent.com/daanishrasheed/DATA607/main/Assignment%201/MTA_Subway_Hourly_Ridership__7_2024.csv", show_col_types = FALSE)

glimpse(dataMta)
```

As mentioned, the target variable is 'ridership', so we need to shorten the dataset to include only the information that has an impact.

```{r}
subset <- dataMta %>%
  select(
    transit_timestamp,
    station_complex,
    borough,
    transit_mode,
    payment_method,
    fare_class_category,
    ridership   # target variable
  )
```
