---
title: "Project 1"
format: html
editor: visual
---

# Chess Tournament Details

## Approach

I would start by cleaning the input file by removing everything that does not constitute the information about any particular player in the tournament: the header of the file, as well as the separator lines. At this point, the file consists of two lines for each player where the first line contains the name, total points, and details of how every single round went for him/her and the second line – his/her state and rating.

I extract the data from every single line using "\|" as a delimiter because of its predictability. The rating of some players is followed by additional comments that I strip using regular expressions to get only the number. I also extract the name of each player's opponent in every round using the same technique and skipping rounds when he/she got a bye or did not participate.

Now I match the player number to his/her rating and calculate the average rating of opponents for each individual player by averaging the ratings of all the players whom he/she played against. Finally, I combine all the required information into one table.

## Code Base

```{r}
library(stringr)
library(dplyr)
data_url <- "https://raw.githubusercontent.com/daanishrasheed/DATA607/refs/heads/main/Project%201/tournamentinfo.txt"

raw_lines <- readLines(data_url)

# Drop blank lines, separator lines (all dashes), and the two header lines
clean_lines <- raw_lines[
  str_trim(raw_lines) != "" &
    !str_detect(str_trim(raw_lines), "^-+$") &
    !str_detect(str_trim(raw_lines), "^(Pair|Num)")
]

length(clean_lines)
```

```{r}
parse1 <- function(x) {
  line1 <- clean_lines[2 * x - 1]
  line2 <- clean_lines[2 * x]
  f1 <- str_trim(str_split(line1, "\\|")[[1]])
  f2 <- str_trim(str_split(line2, "\\|")[[1]])
  
  numpair <- as.integer(f1[1])
  name     <- f1[2]
  points   <- as.numeric(f1[3])
  state <- f2[1]
  pre_rating <- as.integer(str_extract(str_extract(f2[2], "R:\\s*\\d+"), "\\d+"))
  rounded <- f1[4:10]
  
  o <- rounded |>
    str_extract("\\d+") |>
    as.integer()
  o <- o[!is.na(o)]
  
  list(pair = numpair, name = name, state = state,
       points = points, pre_rating = pre_rating, opponents = list(o))
}

n <- length(clean_lines) / 2

players <- lapply(seq_len(n), parse1)

players_df <- bind_rows(lapply(players, function(p) {
  tibble(pair = p$pair, name = p$name, state = p$state,
         points = p$points, pre_rating = p$pre_rating)
}))

head(players_df, 10)
```

\
The above code chunk returns the pair #, name, state, points, and pre-rating for each person in the .txt file.

```{r}
ratings <- setNames(players_df$pre_rating, players_df$pair)

avg_opp <- sapply(players, function(p) {
  opp_ratings <- ratings[as.character(p$opponents[[1]])]
  opp_ratings <- opp_ratings[!is.na(opp_ratings)]
  if (length(opp_ratings) == 0) return(NA_real_)
  round(mean(opp_ratings))
})

players_df$avg_opponent_pre_rating <- avg_opp

head(players_df, 10)
```

To get the average opponents rating for each player, the code builds a lookup table so any rating can be found by a player's number. Then for each player, it grabs their opponents' ratings, drops any missing ones and averages the data that is available.

```{r}
write.csv(players_df, "tournament_ratings.csv", row.names = FALSE)
```

## Conclusion

This parser will take this fixed-width data structure and convert it into a CSV with all of the relevant data formatted nicely in tabular format with the columns, name, state, total score, rating before the tournament and the average rating of their opponents before the tournament, all for each player. The hardest part of doing this was calculating the average of the opponents' rating as I had to find a way to access the pair number of each opponent. I wasn't able to do this with the data as a number, so I had to reformat it as a string and individually extracted each of them.
